How fluid-mechanical erosion creates anisotropic porous media

### 2022-06-27 docs/PhysicsPaper/StokesErosion.tex
NM

Like the permeability, the tortuosity provides a macroscopic perspective
of the porous medium. We define the tortuosity to be the average length
of streamlines passing through the region $[-1,1] \times [-1,1]$. In
particular, for a particular eroded geometry with velocity field $\uu$,
we calculate streamlines $\pp(t)$ that satisfy
\begin{align}
  \dot{\pp}(t) = \uu(\pp), \quad \pp(0) = (-1,y_0),
\end{align}
where $y_0 \in (-1,1)$. The streamlines are calculated with a fourth-order Runge-Kutta method. Then, the length of each streamline, which only depends on its initial $y$-coordinate, is
\begin{align}
  \lambda(y_0) = \int_{0}^{t(y_0)} \|\pp(t)\| \,dt.
\end{align}
The integral limit $t(y_0)$ is chosen so that the $x$-coordinate of
$\pp(t(y_0))$ is $+1$. Then, the longitudinal tortuosity is
\begin{align}
  T_1 = \frac{1}{2}\left(\int_{S}u(-1,y)\lambda(y)\,dy \right)
  \Bigg/
  \left(\int_{S}u(x_0,y)\,dy \right).
  \label{eqn:tortuosity_x}
\end{align} 
\todo[inline]{What is $x_0$ in the denominator? Should that be -1?}
The factor of $1/2$ guarantees that $T_1 \geq 1$, and $T_1 = 1$ if and only if no grains are present. Identical to how we calculate the transverse permeability, we rotate \nick{the configuration of} eroded bodies by 90 degrees and compute a transverse tortuosity $T_2$. Then, another measure of the anisotropy is
\begin{align}
  \anistort = T_2/T_1.
\end{align}
As in the permeability definition of anisotropy, the random configuration of circles used to initialize the erosion simulations will have an anisotropy nearly equal to one. As this configuration erodes, the streamlines in the longitudinal direction are expected to less tortuous than those in the transverse direction, yielding $\anis > 1$.

@@

%---------------------------------------------------------%
\begin{figure*}%[htbp]
\centering
\includegraphics[width = 0.99 \textwidth]{./figs/fig4.pdf}
\caption{
Tortuosity
\label{fig4}
}
\end{figure*}
%---------------------------------------------------------%
% Data from 80-4

As a second way to quantify medium anisotropy, we measure the tortuosity, $T_1$ and $T_2$, in the longitudinal and transverse directions respectively. Figure \ref{fig4}(a), shows how both vary over the course of the simulation, with $T_2$ growing significantly larger than $T_1$, indicating that passive tracers must take more tortuous paths around the partially eroded grains when traveling transverse to the flow.

>>>>

Like the permeability, the tortuosity provides a macroscopic perspective
of the porous medium. We define the tortuosity to be the average length
of streamlines passing through the region $[-1,1] \times [-1,1]$. In
particular, for a particular eroded geometry with velocity field $\uu$,
we calculate streamlines $\pp(t)$ that satisfy
\begin{align}
  \dot{\pp}(t) = \uu(\pp), \quad \pp(0) = (-1,y_0),
\end{align}
where $y_0 \in (-1,1)$. The streamlines are calculated with a fourth-order Runge-Kutta method. Then, the length of each streamline, which only depends on its initial $y$-coordinate, is
\begin{align}
  \lambda(y_0) = \int_{0}^{t(y_0)} \|\pp(t)\| \,dt.
\end{align}
The integral limit $t(y_0)$ is chosen so that the $x$-coordinate of
$\pp(t(y_0))$ is $+1$. Then, the longitudinal tortuosity is
\begin{align}
  T_1 = \frac{1}{2}\left(\int_{S}u(-1,y)\lambda(y)\,dy \right)
  \Bigg/
  \left(\int_{S}u(x_0,y)\,dy \right).
  \label{eqn:tortuosity_x}
\end{align} 
\todo[inline]{What is $x_0$ in the denominator? Should that be -1?}
The factor of $1/2$ guarantees that $T_1 \geq 1$, and $T_1 = 1$ if and only if no grains are present. Identical to how we calculate the transverse permeability, we rotate \nick{the configuration of} eroded bodies by 90 degrees and compute a transverse tortuosity $T_2$. Then, another measure of the anisotropy is
\begin{align}
  \anistort = T_2/T_1.
\end{align}
\todo[inline]{Might eliminate this line so that we don't double-use anisotropy. In any case, I think $(T_2-1)/(T_1-1)$ might be more useful. Will give it thought.}
As in the permeability definition of anisotropy, the random configuration of circles used to initialize the erosion simulations will have an anisotropy nearly equal to one. As this configuration erodes, the streamlines in the longitudinal direction are expected to less tortuous than those in the transverse direction, yielding $\anis > 1$.

@@

%---------------------------------------------------------%
\begin{figure*}%[htbp]
\centering
\includegraphics[width = 0.99 \textwidth]{./figs/fig4.pdf}
\caption{
Tortuosity measurements of the same 80-body simulation featured in Figs.~\ref{fig1} and \ref{fig2}. a) As the bodies erode, the transverse tortuosity $T_2$ exceeds the longitudinal component, indicating more convoluted paths for tracers traveling vertically, transverse to the flow. Also shown are the measurements for the configurations of circles. b) The ratio $(T_2-1)/(T_1-1)$ serves as a counterpart to the anisotropy. Like anisotropy, this ratio grows large over the course of erosion. It can be decomposed into configurational and shape components in a similar manner.
\label{fig4}
}
\end{figure*}
%---------------------------------------------------------%
% Data from 80-4

As a second way to quantify medium anisotropy, we measure the tortuosity, $T_1$ and $T_2$, in the longitudinal and transverse directions respectively. Figure \ref{fig4}(a), shows how both vary over the course of the simulation. As the bodies erode, $T_2$ grows significantly larger than $T_1$, indicating that passive tracers must follow more tortuous paths when traveling transversely and that longitudinal paths are more direct. The ratio $(T_2-1)/(T_1-1)$ provides a second measure of anisotropy to complement the permeability-dependent quantity. We have chosen to subtract one in parenthesis so that both numerator and denominator vanish when no bodies are present, as analogous to resistivity. As seen in Fig.~\ref{fig4}(b), the ratio $(T_2-1)/(T_1-1)$ grows significantly during erosion, up to a peak of nearly 8. This ratio can be decomposed into configurational and shape components, exactly as is done for the anisotropy (i.e.~by replacing a collection of eroded bodies with circles of the same areas and centers of mass). Figure~\ref{fig4}(b) shows that the shape component is greater for the majority of the simulation, but near the end of the simulation, the configurational component is competitive.
